Towards Practical First-Order Model Counting

### 2023-10-10 paper.tex
A high-level process overview figure

\usepackage{listings}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{microtype}
\usepackage{amsfonts}
\usepackage{mathtools}

@@

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem*{remark}{Remark}

@@

\crefname{line}{line}{lines}

\DeclareMathOperator{\Dom}{Dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Preds}{Preds}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator*{\argmin}{argmin}

@@

\paragraph{Contributions.}
\begin{itemize}
  \item Converting the recursive equations into a C++ program, which can then be
        compiled and executed to obtain numerical values (see \cref{sec:cpp}).
  \item Support for infinite precision integers using the GNU Multiple Precision
        Arithmetic Library.
\end{itemize}

\section{Snippets}

\begin{itemize}
  \item The evaluation of base cases is done by simplifying the clauses and then
        using \textsc{Crane} to find the base cases. First, while traversing the
        graph to find the equations, we store two maps objects:
        \texttt{clause\_func\_map} (which stores the mapping from the function
        names to the formulae, whose model count they represent) and
        \texttt{var\_domain\_map} (which stores the mapping from the variable
        names to the domains whose sizes they represent). Then, a particular
        domain is selected (using the algorithm described in previous reports),
        and the clauses are simplified. Then, \textsc{Crane} is called on those
        clauses to evaluate the base cases. After that, we change the function
        names and variable to make it consistent with the previous domain to
        variable mapping, and append these base cases to the set of equations.
  \item Finding a sufficient set of base cases. We know that if, say, on the RHS
        of all equations, the domain size appears as
        $m - c_1, m - c_2, \dots, m - c_k$, then finding
        $f(0, x_1, x_2, \dots)$, $f(1, x_1, x_2, \dots)$,
        $\dots f(m_0, x_1, x_2, \dots)$ for every function $f$, where
        $m_0 = \max(c_1, c_2, \dots c_k) - 1$ forms a sufficient set of base
        cases. Hence, in order to do the same efficiently, we can take that
        domain for which $m_0$ is the minimum, i.e.
        $\argmin(\max(c_1, c_2, \dots c_k))$.
\end{itemize}

\subsection{Bar}

The previous method of base case evaluation on setting a domain to size zero or
one had the following error. In case a domain size was set to zero, it assumed
that those predicates which were deleted from the clauses could take any truth
value over the entire domain and the rest were fully covered by the remaining
clauses.

For example, consider the formula
\begin{equation}\label{eq:example}
  \begin{split}
    \forall x \in \Delta \forall y, z \in \Gamma &: P(x) \lor Q(y, z) \\
    \forall y \in \Gamma^\top &: Q(y, z).
  \end{split}
\end{equation}
In this case, if we set $|\Delta|$ to zero, the transformed formula would be
\[
  \forall y, z \in \Gamma^\top : Q(y, z),
\]
and the set of removed predicates is empty, and hence the model count returned
would be $1$. However, the actual model count should be
$2^{|\Gamma|^2 - |\Gamma^\top|^2}$.

We solve this problem by converting clauses with universal quantifiers over the
empty domain to tautologies, hence retaining all the predicates that have no
argument assigned to the empty domain. For example, we would convert the above
mentioned formula to
\begin{align*}
  \forall y, z \in \Gamma &: Q(y, z) \lor \neg Q(y, z) \\
  \forall y, z \in \Gamma^\top &: Q(y, z)
\end{align*}
The model count returned by this will also consider the truth value of $Q$ over
$y \not \in \Gamma^\top$ or $z \not \in \Gamma^\top$.

@@

TODO: Introduce terms: left-hand side (LHS), right-hand side (RHS).

@@

\begin{remark}
  The LHS of an equation is always a signature.
\end{remark}

For any signature or clause $C$, argument or variable $x$, and number or
constant $t$, we shall write $C[t / x]$ for the result of substituting $t$ for
all occurrences of $x$ in $C$.
% TODO: check if I'm still using this notation in all these cases
% TODO: explain how this works for function calls too

\section{Identifying a Sufficient Set of Base Cases}

\begin{algorithm}[t]
  \caption{Identifying a set of sufficient base cases}\label{alg1}
  \KwIn{set $D$ of dependencies}
  \KwOut{set $B$ of base cases}

  $B \gets \emptyset$\;
  \ForEach{$(f(\mathbf{x}), g(\mathbf{y} - \mathbf{c})) \in D$}{
    \ForEach{$c_{i} \in \mathbf{c}$}{
      \For{$n \gets 0$ \KwTo $c_{i} - 1$}{\label{line:lim}
        $B \gets B \cup \{\, f(\mathbf{x})[x_{i}/n] \,\}$\;\label{line:insert}
        \lIf{$f \ne g$}{$B \gets B \cup \{\, g(\mathbf{y})[y_{i}/n] \,\}$}\label{line:signature}
      }
    }
  }
\end{algorithm}

The following steps were followed while finding the base cases:
\begin{enumerate}
  \item Expand the summations in each equation. Here we expand the summations of
        the form:
        $\sum_{x=0}^{x_{1}}<\textnormal{something}> \cdot [a \le x < b]$, or
        similar inequalities where x is bounded by constants and $a$ and $b$ are
        constants, by substituting the value of $x$ from $a$ to $b-1$. For
        example, we replace
        $\sum_{x=0}^{x_1} \binom{x_{1}}{x} f(x_1 - x) \cdot [0 \le x < 2]$ by
        $\binom{x_{1}}{0} f(x_1) + \binom{x_{1}}{1} f(x_1-1)$.
  \item Next, we find the dependencies of those functions that appear on the
        RHS of any equation. Consider, for example the following set
        of equations:
        \begin{align*}
          f_{0}(m, n) &= f_{1}(m-1, n) + f_{2}(m, n-1)\\
          f_{1}(m, n) &= f_{1}(m-1, n-1) \times f_{2}(m-2, n-1)\\
          f_{2}(m, n) &= 2 \times f_{1}(m-3, n-1)
        \end{align*}
        In this case, the dependencies computed are
        \begin{align*}
          f_{1}(m, n) &\mapsto \{\, f_{1}(m-1, n-1), f_{2}(m, n-1) \,\}\\
          f_{2}(m, n) &\mapsto \{\, f_{1}(m-3, n-1) \,\}
        \end{align*}
  \item Now, we find a domain that has only terms of the form $x-1$ appearing on
        the RHS of the dependencies. The base cases are then calculated by
        setting this domain size to zero. For the above example, $n$ is the
        selected domain, and not $m$ since there are $m-2$ and $m-3$ terms
        appearing in the arguments.
\end{enumerate}

\paragraph{Limitations of the current implementation.} Ideally, we should
calculate the base cases by finding the base cases up to
$\max(c_{1}, c_{2}, \dots) - 1$. However, currently only empty and singleton
domains are supported.

\begin{itemize}
  \item \cref{line:lim}: the limit is 2 for the arg $(x-3)$.
  \item \cref{line:insert} for $f(\mathbf{x}) = f(x_{1}, x_{2})$,
        $\text{arg} = x$, and $n = 0$, add $f(0, x_{2})$ to $B$.
  \item note that on \cref{line:signature} it is the signature and not the
        original function call.
\end{itemize}

The algorithm is described as \cref{alg1}. Here, a dependency is a pair
$(a, b)$, where $a$ is the signature on the LHS of each equation and $b$ is each
function call on the RHS of the equation. For example, for the equations
\begin{align*}
  f(m, n) &= g(m-1, n) + f(m-2, n-1)\\
  g(m, n) &= f(m-1, n-2) + g(m-1, n-1)
\end{align*}
the dependencies are
\begin{multline*}
  (f(m, n), g(m - 1, n)), (f(m, n), f(m - 2, n - 1)),\\
  (g(m, n), f(m - 1, n - 2), g(m - 1, n - 1)).
\end{multline*}

\section{Algorithm to Transform CNF Based on Domain Sizes}

\begin{algorithm}[t]
  \caption{Transforming Formulas Based on Domain Sizes}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, target domain size $n \in \{\, 0, 1 \,\}$, domain size function $|\cdot|$}
  \KwOut{formula $\phi'$}
  $P^{-} \gets \emptyset$; $P^{+} \gets \emptyset$; $\phi' \gets \emptyset$\;
  \uIf{$|\Delta| = 0$}{
    \ForEach{clause $C \in \phi$}{
      \uIf{$\Delta \in \Doms(C)$}{
        $P^{-} \gets P^{-} \cup \Preds(C)$\;
      }
      \Else{
        $P^{+} \gets P^{+} \cup \Preds(C)$\;
        $\phi' \gets \phi' \cup \{\, C \,\}$\;
      }
    }
    $P^{-} \gets P^{-} \setminus P^{+}$\;
  }
  \ElseIf{$|\Delta| = 1$}{
    $c \gets$ a new constant symbol\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[c / v]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}
% TODO: the removed empty domains should be replaced by smoothing/tautological
% constructions

% TODO: either the target domain size or the domain size function needs to go

\begin{itemize}
  \item Formulas are in CNF. Formulas are (multi) sets of clauses.
  \item $P^{-}$ --- removed predicates
  \item $P^{+}$ --- retained predicates
  \item $\phi'$ --- transformed version of $\phi$
  \item $C'$ --- transformed version of clause $C$
  \item For any clause $C$, let $\Doms(C)$, $\Preds(C)$, and $\Vars(C)$ denote
        respectively the set of domains, predicates, and variables in $C$. For
        any variable $v$, let $\Dom(v)$ denote its domain. Note that, for any
        clause $C$, we have that
        $\Doms(C) = \{\, \Dom(v) \mid v \in \Vars(C) \,\}$. Similarly, for any
        predicate $P$, let $\Doms(P)$ denote the set of domains associated with
        $P$.
\end{itemize}

@@

@@

The target is to generate C++ code that can evaluate numerical values of the
model counts based on the equations generated by \textsc{Crane}. There are two
ways to do the same.
\begin{enumerate}
  \item Generate C++ code by traversing the FCG, similar to what is done in
        \texttt{OutputVisitor.scala}.
  \item Parse the equations generated by \textsc{Crane} after simplifying in
        wolfram and then generate C++ code.
\end{enumerate}

The problem with the first approach is that, while generating base cases, the
subsequent calls to \textsc{Crane} do not necessarily have the same meanings for
the function arguments and the functions. For example, if
$f_{1}(x_{0}, x_{1}, x_{2})$ represents the model count of a constrained formula
$\phi$, where $f_{1}$ is an auxiliary formula and $x_{0} = |A|$, $x_{1} = |B|$,
$x_{2} = |C|$, and $A$, $B$, $C$ are domains and we want to evaluate
$f_{1}(0, x_{1}, x_{2})$ (i.e., set $|A| = 0$), then \textsc{Crane} may return
the required model count as $f_{0}(x_{0}, x_{1})$, where $x_{0} = |B|$ and
$x_{1} = |C|$. We will then need to translate this to $f_{1}(0, x_{1}, x_{2})$.
Also, the second approach can be done in linear time in the length of the
formula using
the
\href{https://en.wikipedia.org/wiki/Shunting_yard_algorithm#:~:text=In%20computer%20science%2C%20the%20shunting,abstract%20syntax%20tree%20(AST).}{Shunting
  Yard Algorithm}. Hence, we stick to the second approach.

\subsection{Our Approach}

@@

The RHS of each equation in $E$ is used to define the body of the equation
corresponding to the LHS of that equation. The function body (for a function
\texttt{func} corresponding to equation $e$) is formed as follows.
\begin{enumerate}
  \item First, we check if the evaluation is already present in the cache. If
        so, then we return the cache element. The cache accesses are done using
        the \texttt{get\_elem} function (definition given in the example), which
        resizes the cache if the accessed index is out of range.
  \item If the element is absent, then we decide if the arguments corresponding
        to $e$ or one of the functions corresponding to the base cases, based on
        the value of the arguments. If it corresponds to the base cases, then we
        directly call the base case function and return its value. Else, we
        evaluate the value using the RHS, store the evaluated value in the cache
        and return the evaluated value. Note that in this step, we only call the
        base case function with one more constant argument that \texttt{func}.
        For example, \texttt{f0(x, y)} would call \texttt{f0\_0x(y)} if
        \texttt{x == 0} and \texttt{f0\_x0(x)} if \texttt{y == 0}.
  \item In order to translate the RHS, we convert $\sum_{x=a}^{b} \texttt{exp}$
        to
\begin{lstlisting}
([y,z,...](){
    int sum = 0;
    for(int x = a; x <= b; x++)
        sum += exp;
    return sum;
})()
\end{lstlisting}
        where $y, z, \dots$ are the free variables present in \texttt{exp}.
\end{enumerate}

>>>>

\usepackage{listings}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{microtype}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{tikz}

@@

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem*{remark}{Remark}
\newtheorem{fact}{Fact}

@@

\crefname{line}{line}{lines}
\crefalias{formula}{equation}
\crefname{formula}{Clause}{Clauses}
\creflabelformat{formula}{#2\textup{(#1)}#3}

\DeclareMathOperator{\Dom}{Dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Preds}{Preds}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand{\expr}{\mathtt{expr}}

\SetKwFunction{Crane}{Crane}
\SetKwFunction{Propagate}{Propagate}
\SetKwFunction{FindBaseCases}{FindBaseCases}

\usetikzlibrary{arrows.meta}
\usetikzlibrary{positioning}
\usetikzlibrary{shapes}
\usetikzlibrary{fit}

@@

\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \node at (0, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (alg1) {\Cref{alg:main}};
    \node[draw,rounded rectangle] at (6, 0) (simplification) {Simplification};
    \node[draw,rounded rectangle] at (9, 0) (compilation) {Compilation};

    \node[draw,rounded rectangle,dashed] at (12, 0) (cpp) {C++ code};
    \node at (12, -1) (sizes) {Domain sizes};

    \node at (15, 0) (count) {Model count};

    \node[draw,rounded rectangle] at (3, -2) (alg2) {\Cref{alg1}};
    \node[draw,rounded rectangle,left = 0.1cm of alg2] (crane) {\textsc{Crane}};
    \node[draw,rounded rectangle,right = 0.1cm of alg2] (alg3) {\Cref{alg2}};

    \node[draw,fit={(alg1) (simplification) (compilation) (crane) (alg2) (alg3)},inner ysep=7pt,yshift=5pt] {};
    \node at (1.1, 0.5) {\textsc{Crane2}};

    \draw[->] (formula) -- (alg1);
    \draw[->] (alg1) -- node[above] {$E$} (simplification);
    \draw[->] (simplification) -- node[above] {$E'$} (compilation);
    \draw[->] (compilation) -- (cpp);
    \draw[->] (sizes) -- (cpp);
    \draw[->] (cpp) -- (count);

    \draw[->,dashed] (alg1) -- node[midway,left] {uses} (crane);
    \draw[->,dashed] (alg1) -- node[midway] {uses} (alg2);
    \draw[->,dashed] (alg1) -- node[midway,right] {uses} (alg3);
  \end{tikzpicture}
  \caption{Using \textsc{Crane2} to compute the model count of formula $\phi$.\
    The formula is compiled into a set of equations $E$, which are then
    algebraically simplified and compiled to a C++ program. This program can
    then be run with different command line arguments to compute the model count
    of $\phi$ for various domain sizes.}
\end{figure*}

The evaluation of base cases is done by simplifying the clauses and then using
\textsc{Crane} to find the base cases. First, while traversing the graph to find
the equations, we store two maps: $\mathcal{F}$ (which stores the mapping from
the function names to the formulae, whose model count they represent) and
$\mathcal{D}$ (which stores the mapping from the variable names to the domains
whose sizes they represent). Then, a particular domain is selected (using the
algorithm described in previous reports), and the clauses are simplified. Then,
\textsc{Crane} is called on those clauses to evaluate the base cases. After
that, we change the function names and variable to make it consistent with the
previous domain to variable mapping, and append these base cases to the set of
equations.

\paragraph{Contributions.}
\begin{itemize}
  \item \Cref{sec:identifying}
  \item \Cref{sec:simplifying}
  \item Converting the recursive equations into a C++ program, which can then be
        compiled and executed to obtain numerical values (see \cref{sec:cpp}).
  \item Support for infinite precision integers using the GNU Multiple Precision
        Arithmetic Library.
  \item The experiments are (going to be) more comprehensive than any other
        experimental study performed on (W)FOMC algorithms (TODO: references to
        my previous paper and some of the \textsc{ForcLift} and
        \textsc{FastWFOMC} papers that contain experimental results).
\end{itemize}

\Cref{sec:identifying,sec:cpp} deal with algebraic constructs whereas
\cref{sec:simplifying} deals with logic.

@@

\subsection{Algebra}

\paragraph{Notation.}
We write $\expr{}$ for an arbitrary algebraic expression. For any signature or
clause $C$, argument or variable $x$, and number or constant $t$, we shall write
$C[t / x]$ for the result of substituting $t$ for all occurrences of $x$ in $C$.
% TODO: check if I'm still using this notation in all these cases
% TODO: explain how this works for function calls too

@@

\begin{definition}
  An \emph{equation} is always of the form $f(\mathbf{x}) = \expr{}$, where
  $f(\mathbf{x})$ is a signature, and $\expr{}$ is an algebraic expression.
  Henceforth, we call $f(\mathbf{x})$ and $\expr{}$ the left-hand side (LHS) and
  the right-hand side (RHS) of the equation, respectively.
\end{definition}
% TODO: give some examples of formulas that: 1) introduce new variables, 2) call
% the function itself, 3) call some other function, 4) is more nested than just
% a sum.

\begin{definition}
  A \emph{base case} is a function call where all arguments are either variables
  or constants (i.e., there is no subtraction within any argument), and there is
  at least one constant.
\end{definition}

\subsection{Logic}

% TODO: what notation am I using for FOL? Colon after every quantifier?

% TODO: (Similar to CNF, maybe QBF). Incomplete. Need (W)FOMC semantics. What is
% an interpretation? What is a model?

A \emph{formula} is a set of clauses. A \emph{clause} is of the form
$\forall x_{1} \in \Delta_{1}\forall x_{2} \in \Delta_{2}\dots\forall x_{n} \in \Delta_{n} \phi(x_{1}, x_{2}, \dots, x_{n})$,
where $\phi$ is a disjunction of literals that only contain variables
$x_{1}, \dots, x_{n}$ (and any constants). A \emph{literal} is either an atom or
its negation. An \emph{atom} is of the form $P(t_{1}, \dots, t_{m})$, where $P$
is a predicate, and $t_{1}, \dots, t_{m}$ are terms. A \emph{term} is either a
variable or a constant.
% TODO: explain set-based notation for a clause (iterating over all literals,
% maybe I should have a Lits() operator? No!!!)

For any clause $C$, let $\Preds(C)$ denote the set of predicates in $C$. For any
clause or literal $l$, let $\Vars(l)$ denote the set of variables in $l$.

For any (bound) variable $v$, let $\Dom(v)$ denote the domain over which $v$ is
quantified. For a literal $l$,
$\Doms(l) \coloneqq \{\, \Dom(v) \mid v \in \Vars(l) \,\}$. Similarly, for any
clause $C$, let $\Doms(C) = \{\, \Dom(v) \mid v \in \Vars(C) \,\}$.

\section{The Main Algorithm: Finding the Definitions of the Base Cases}

\begin{algorithm}[t]
  \caption{Finding the definitions of the base cases}\label{alg:main}
  \KwIn{formula $\phi$}
  \KwOut{a set $E$ of equations that define $B$}
  $(E, \mathcal{F}, \mathcal{D}) \gets \Crane{$\phi$}$\;
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$E$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i}$ is a constant}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $(E', \textunderscore, \textunderscore) \gets \Crane{$\psi$}$\;
    $E \gets E \cup E'$\;
  }
\end{algorithm}

See \cref{alg:main}

% TODO: extend it to work with base cases that are themselves recursive

% TODO: what does it mean for a set of equations to define a set of base cases?

% TODO: for base cases, we will also use the f(\mathbf{x}) notation

\section{Identifying a Sufficient Set of Base Cases}\label{sec:identifying}

\begin{algorithm}[t]
  \caption{\protect\FindBaseCases{$E$}}\label{alg1}
  \KwIn{set $E$ of equations}
  \KwOut{set $B$ of base cases}

  $B \gets \emptyset$\;
  \ForEach{equation $(f(\mathbf{x}) = \expr{}) \in E$}{
    \ForEach{function call $g(\mathbf{y} - \mathbf{c}) \in \expr{}$}{\label{line:functioncall}
      \ForEach{$c_{i} \in \mathbf{c}$}{
        \For{$n \gets 0$ \KwTo $c_{i} - 1$}{\label{line:lim}
          $B \gets B \cup \{\, f(\mathbf{x})[x_{i}/n] \,\}$\;\label{line:insert}
          \lIf{$f \ne g$}{$B \gets B \cup \{\, g(\mathbf{y})[y_{i}/n] \,\}$}\label{line:signature}
        }
      }
    }
  }
\end{algorithm}

We know that if, say, on the RHS of all equations, the domain size appears as
$m - c_1, m - c_2, \dots, m - c_k$, then finding $f(0, x_1, x_2, \dots)$,
$f(1, x_1, x_2, \dots)$, $\dots f(m_0, x_1, x_2, \dots)$ for every function $f$,
where $m_0 = \max(c_1, c_2, \dots c_k) - 1$ forms a sufficient set of base
cases. Hence, in order to do the same efficiently, we can take that domain for
which $m_0$ is the minimum, i.e. $\argmin(\max(c_1, c_2, \dots c_k))$. Ideally,
we should calculate the base cases by finding the base cases up to
$\max(c_{1}, c_{2}, \dots) - 1$. However, currently only empty and singleton
domains are supported.

\begin{itemize}
  \item \cref{line:functioncall} iterates over all function calls in the
        algebraic expression.
  \item note that on \cref{line:signature} it is the signature and not the
        original function call.
\end{itemize}

% TODO: remove the 'dependencies' data structure from the description

The following steps were followed while finding the base cases:

First, expand the summations in each equation. Here we expand the summations of
the form: $\sum_{x=0}^{x_{1}} \expr{} \cdot [a \le x < b]$ or similar
inequalities where $x$ is bounded by constants and $a$ and $b$ are constants, by
substituting the value of $x$ from $a$ to $b-1$. For example, we replace
$\sum_{x=0}^{x_1} \binom{x_{1}}{x} f(x_1 - x) \cdot [0 \le x < 2]$ by
$\binom{x_{1}}{0} f(x_1) + \binom{x_{1}}{1} f(x_1-1)$.

Now, we find a domain that has only terms of the form $x-1$ appearing on the RHS
of the dependencies. The base cases are then calculated by setting this domain
size to zero. For the above example, $n$ is the selected domain, and not $m$
since there are $m-2$ and $m-3$ terms appearing in the arguments.

The algorithm is described as \cref{alg1}.

\begin{example}
  TODO: add an example where \cref{alg1} runs on either
  \begin{align*}
    f_{0}(m, n) &= f_{1}(m-1, n) + f_{2}(m, n-1)\\
    f_{1}(m, n) &= f_{1}(m-1, n-1) \times f_{2}(m-2, n-1)\\
    f_{2}(m, n) &= 2 \times f_{1}(m-3, n-1)
  \end{align*}
  or
  \begin{align*}
    f(m, n) &= g(m-1, n) + f(m-2, n-1)\\
    g(m, n) &= f(m-1, n-2) + g(m-1, n-1).
  \end{align*}

  \begin{itemize}
    \item \cref{line:lim}: the limit is 2 for the arg $(x-3)$.
    \item \cref{line:insert} for $f(\mathbf{x}) = f(y, z)$, $x_{i} = y$, and
          $n = 0$, add $f(y, z)[y/0] = f(0, z)$ to $B$.
  \end{itemize}
\end{example}

\section{Propagating Domain Size Assumptions}\label{sec:simplifying}

\subsection{Motivation: Why Tautologies Are Needed and Simply Removing The
  Clauses Does Not Work}

\begin{fact}
  Assuming that domain $\Delta$ is empty, any clause that contains
  `$\forall x \in \Delta$' (for any variable $x$) is vacuously satisfied by all
  interpretations.
\end{fact}

For example, consider the formula
\begin{align}
  \forall x \in \Delta \forall y, z \in \Gamma &: P(x) \lor Q(y, z)\label[formula]{eq:example1}\\
  \forall y, z \in \Gamma' &: Q(y, z)\label[formula]{eq:example2}
\end{align}
and assume that $\Gamma' \subseteq \Gamma$. In this case, if we set $|\Delta|$
to zero and remove clauses with variables quantified over $\Delta$, we get
\begin{equation}\label[formula]{eq:simplified}
  \forall y, z \in \Gamma' : Q(y, z),
\end{equation}
but the model count of \cref{eq:simplified} is one. However, the actual model
count should be $2^{|\Gamma|^2 - |\Gamma'|^2}$. That is, $Q$ as a relation is a
subset of $\Gamma \times \Gamma$. While \cref{eq:example1} becomes vacuously
true, \cref{eq:example2} fixes the value of $Q$ over
$\Gamma' \times \Gamma' \subseteq \Gamma \times \Gamma$. Hence, the number of
different values that $Q$ can take is
$|(\Gamma \times \Gamma) \setminus (\Gamma' \times \Gamma')| = |\Gamma|^{2} - |\Gamma'|^{2}$.

We address this issue by converting clauses with universal quantifiers over the
empty domain to tautologies, hence retaining all the predicates that have no
argument assigned to the empty domain. For example, we would convert
\cref{eq:example1,eq:example2} to
\begin{align*}
  \forall y, z \in \Gamma &: Q(y, z) \lor \neg Q(y, z) \\
  \forall y, z \in \Gamma' &: Q(y, z).
\end{align*}
The model count returned by this will also consider the truth value of $Q$ over
$y \in \Gamma \setminus \Gamma'$ or $z \in \Gamma \setminus \Gamma'$.

\subsection{The Solution}

\begin{algorithm}[t]
  \caption{\protect\Propagate{$\phi$, $\Delta$, $n$}}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, domain size $n \in \{\, 0, 1 \,\}$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \uIf{$n = 0$}{
    \ForEach{clause $C \in \phi$}{
      \lIf{$\Delta \not\in \Doms(C)$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      $C' \gets \{\, l \in C \mid \Delta \not\in \Doms(l) \,\}$\;
      \If{$\Delta \in \Doms(C)$ {\bf and} $C' \ne \emptyset$}{
        $l \gets \text{an arbitrary literal in } C'$\;
        $\phi' \gets \phi' \cup \{\, C' \cup \{\, \neg l \,\} \,\} $\;
      }% TODO: find the source code and check: 1) what we're doing with unit clauses, 2) whether we just randomly pick any literal to negate
    }
  }
  \Else{
    $c \gets$ a new constant symbol\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[c / v]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}
% TODO: it would be easy to extend the second part to an arbitrary constant

@@

\begin{itemize}
  \item Similar to generalised domain recursion (what's the difference?).
  \item Clauses where all literals have variables quantified over the empty
        domain are still removed.
  \item $C'$ is guaranteed to be non-empty.
\end{itemize}

@@

The target is to generate C++ code that can evaluate numerical values of the
model counts based on the equations generated by \textsc{Crane}. We achieve this
by parsing the equations generated by \textsc{Crane}, simplifying them, and then
generating C++ code. This approach can be done in linear time in the length of
the formula using
the
\href{https://en.wikipedia.org/wiki/Shunting_yard_algorithm#:~:text=In%20computer%20science%2C%20the%20shunting,abstract%20syntax%20tree%20(AST).}{Shunting
  Yard Algorithm}.

@@

The RHS of each equation in $E$ is used to define the body of the equation
corresponding to the LHS of that equation. The function body (for a function
\texttt{func} corresponding to equation $e$) is formed as follows.

First, we check if the evaluation is already present in the cache. If so, then
we return the cache element. The cache accesses are done using the
\texttt{get\_elem} function (definition given in the example), which resizes the
cache if the accessed index is out of range.

Second, if the element is absent, then we decide if the arguments corresponding
to $e$ or one of the functions corresponding to the base cases, based on the
value of the arguments. If it corresponds to the base cases, then we directly
call the base case function and return its value. Else, we evaluate the value
using the RHS, store the evaluated value in the cache and return the evaluated
value. Note that in this step, we only call the base case function with one more
constant argument than \texttt{func}. For example, \texttt{f0(x, y)} would call
\texttt{f0\_0x(y)} if $x = 0$ and \texttt{f0\_x0(x)} if $y = 0$.

Third, to translate the RHS, we convert $\sum_{x=a}^{b} \expr{}$ to
\begin{lstlisting}[escapeinside={(*}{*)}]
([y,z,...](){
    int sum = 0;
    for(int x = a; x <= b; x++)
        sum += (*$\expr{}$*);
    return sum;
})()
\end{lstlisting}
where $y, z, \dots$ are the free variables present in $\expr{}$.

\section{Experiments}

\section{Conclusion}
